---
title: "Categorical Data"
author: "Ted Laderas"
date: "9/28/2017"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(learnr)
library(gridExtra)
```

## What is Exploratory Data Analysis?

- Pioneered by Tukey
- Detective work on your data
- An attitude towards data, not just techniques
- 'Find patterns, reveal structure, and make tenative model assessments (Behrens)'

### Why EDA and Visualization?

+ We need to become aware of patterns in the data.
+ Sometimes these patterns are desirable:
    + Associations
+ Sometimes they are not desirable:
    + Experimental Artefacts

Visualization is good because we are good at evaluating things visually.

### Good EDA visualizations tell us something about the data

+ Is there an association between smoking and death?
+ Is this association due to something related to smoking?
    + What are some examples?

## What are important associations?

Let's think about this. 

If we know about whether someone smokes, do we know whether they are more likely to die?

Compare it to whether somone wears green. Does this give us any information about whether they are more likely to die?

```{r letter-a, echo=FALSE}
question("If someone wears green are they more likely to die?",
  answer("Yes. Because Lucky the Leprechaun will be angry.", message="Not correct."),
  answer("No. The two variables are not associated", correct=TRUE, message="Yes, that's correct!"),
  allow_retry = TRUE
)
```

## The Data (Whickham)

We're going to look at some data of from a group study of 1314 female smokers. It's called the `Whickham` dataset. 

Let's take a look at a summary of the data. 

```{r message=FALSE, warning=FALSE}
library(mosaic)
data(Whickham)
library(tidyverse)

summary(Whickham)
```

```{r summary, echo=FALSE}
question("What can we learn from this table?",
  answer("There are more alive people than dead people in the data.", correct=TRUE, message="Correct. There are more alive people (945) than dead people (369) in this dataset."),
  answer("There are two variables", message="Not quite. There's `outcome`, `smoker`, and `age`."),
  answer("You are either considered either a smoker or a non-smoker", correct=TRUE, message="Yes, that's correct! This dataset doesn't account for how long you smoked."),
  answer("The age range of the patients includes minors.", message="Incorrect. Look at the min value for age."),
  answer("The average age of the participant was around 47 years of age", correct="Yes, the mean age was 46.92 years of age. Yes, you can see it as the mean value for the age" ),
  allow_retry = TRUE
)
```

## Let's look at our Outcome

Our outcome in our dataset is death. One tool we often use to explore the data is the 

```{r}
table(Whickham$outcome)
```

Here's another way to look at the data, a bar plot. You can instantly see which of the groups is larger.

```{r}
Whickham %>% ggplot(aes(x=outcome)) + geom_bar()
```

### Now let's look at Smoking Status

One way of looking at the data is to make a summary table. 

```{r}
table(Whickham$smoker)
```

Here's the bar plot: 

```{r}
Whickham %>% ggplot(aes(x=smoker)) + geom_bar()
```

## Two Variables Walk Into A Bar

Now that we've seen the variables by themselves, we can consider whether these two variables are associated.

Try to phrase the association in terms of the variables, such as "If you smoke, are you more likely to die?"

### Perfect Association

Before we start assessing whether our two variables are associated, let's think about what we're looking for.

One case is that `smoking` would perfectly predict `outcome`. That is, if you smoke, you would die. The association would look like this:

```{r}
tab <- data.frame(outcome=factor(c("Alive", "Dead", "Alive", "Dead"), levels=c("Alive", "Dead")), smoker=factor(c("No", "No", "Yes", "Yes"), levels=c("Yes", "No")), value=c(500, 10, 7, 300), key=c("A", "B","C", "D"))

vis1 <- tab %>% ggplot(aes(x=outcome, y=smoker)) + geom_tile(fill="white", colour="black") + 
  geom_text(aes(label=value))

vis2 <- tab %>% ggplot(aes(x=outcome, fill=smoker, y=value)) + geom_bar(color="black",stat="identity") +
  geom_text(aes(label=key))

lay <- t(as.matrix(c(1,2), nrow=1))
grid.arrange(vis1, vis2, layout_matrix=lay)
```

### What if there is no association between smoking and death?

This is harder to detect with two by two tables, but it is easier to detect with the proportional plot.

```{r}
tabNoAssoc <- data.frame(outcome=factor(c("Alive", "Dead", "Alive", "Dead"), levels=c("Alive", "Dead")), smoker=factor(c("No", "No", "Yes", "Yes"), levels=c("Yes", "No")), value=c(250, 100, 250, 100), key=c("A", "B","C", "D"))

vis1 <- tabNoAssoc %>% ggplot(aes(x=outcome, y=smoker)) + geom_tile(fill="white", colour="black") + 
  geom_text(aes(label=value))

vis2 <- tabNoAssoc %>% ggplot(aes(x=outcome, fill=smoker, y=value)) + geom_bar(color="black",stat="identity") +
  geom_text(aes(label=key))

lay <- t(as.matrix(c(1,2), nrow=1))
grid.arrange(vis1, vis2, layout_matrix=lay)
```

```{r letter-b, echo=FALSE}
```


## The 2x2 table

So let's look at how these two variables really predict each other.

```{r}
testTab <- table(Whickham$outcome, Whickham$smoker) %>% as.data.frame() 
colnames(testTab) <- c("outcome", "smoker", "Freq")

testTab %>% ggplot(aes(x=outcome, y=smoker)) + geom_tile(fill="white", colour="black") + 
  geom_text(aes(label=Freq))
```

```{r}
Whickham %>% ggplot(aes(x=outcome, fill=smoker)) + geom_bar(color="black")
```

```{r}
Whickham %>% ggplot(aes(x=outcome, fill=smoker)) + geom_bar(position="fill", color="black")
```

```{r letter-c, echo=FALSE}
question("Based on this plot, if you smoke will you be more likely to be alive or dead?",
  answer("Dead, because that's what makes common sense.", message="Not correct."),
  answer("Alive, because that's what the data tells us", correct=TRUE, message="Yes, that's correct!")
)
```

## The effect of Age on the Data

Let's assess what happens with number of deaths as we look at older patients.

Adjust the slider on the left and see what happens to the number of deaths as we increase the average age of the group.

```{r}
sliderInput("slidy", "Age Cutoff", min=min(Whickham$age)+1, max=max(Whickham$age)-1, value=c(30, 83),step = 5)
plotOutput("distPlot")
```

```{r context="server"}
  output$distPlot <- renderPlot(
      exMelt() %>% ggplot(aes(x=status, fill=outcome)) + geom_bar(position="fill", color="black") #+
        #facet_wrap(facets = ~AgeCat)
      )
    
    exMelt <- reactive({
      cutoff = min(input$slidy)
      Whickham %>% filter(age > cutoff) %>% mutate(status="allPatients")
        #mutate(AgeCat=factor(ifelse(age > cutoff, "older", NA))) 
    })
```

```{r letter-d, echo=FALSE}
question("What do you notice about the number of deaths as you increase the age of the subgroup?",
  answer("The number of deaths decreases, which means that as you get older, you are less likely to die.", message="Not correct."),
  answer("The number of deaths increases, which means that as you get older, you are more likely to die.", correct=TRUE, message="Yes, that's correct!")
)
```

## Not as easy as we thought!

So as you get older, you're more likely to die. This may be messing up our overall results! 

Let's ask the question again, with a younger group: are smokers under 44 more likely to die than non smokers?

```{r}
sliderInput("slidy2", "Age Cutoff", min=min(Whickham$age)+1, max=max(Whickham$age)-1, value=c(18, 83),step = 2)
plotOutput("distPlot2")
```

```{r context="server"}
  output$distPlot2 <- renderPlot(
      exMelt() %>% ggplot(aes(x=smoker, fill=outcome)) + geom_bar(position="fill", color="black")
      )
    
    exMelt <- reactive({
      cutoffMax = max(input$slidy2)
      cutoffMin = min(input$slidy2)
      Whickham %>% filter(age < cutoffMax & age > cutoffMin) %>% mutate(status="allPatients")
    })
```

At what age group are the smoking and 

How does this change our research question?

Is smoking associated with death for those under 44 years of age?

## Your Turn

We're going to look at another dataset called NHANES (National Health and Nutrition Examination Survey). 

We're going to 
